### 2019-06-06 2-Chapters/2-Chapter/chapter2.tex
Lots of small changes
- added a title/author etc in metadata
- cleaner small proof with an environment

% \TODOL{Je dois unifier les notations, et utiliser $Y_{k,t}$ pour les flux qui sont iid et $r(t)$ pour les récompenses effectivement observées (comme dans les chapitres 5 et 6)}

@@

\begin{definition}[Kullback-Leibler divergence]\label{def:2:KLDivergence}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
    The KL divergence between two distributions $d_1$ and $d_2$ is defined as
    \begin{align}\label{eq:2:Kullback-LeiblerDivergenceExpFamily1}
        \KL\left(d_1, d_2\right) = \int d_1 \log\left(\frac{d_2}{d_1}\right) \lambda(\mathrm{d}x) = \E_{d_1} \left[ \log\left(\frac{d_2}{d_1}\right) \right] \in \mathbb{R}\cup\{\pm\infty\}.
    \end{align}
\vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar defnbar, comment if needed
\end{definition}

@@

\label{par:2:notationsExponentialFamiliesBernoulliGaussian}
% \begin{itemize}
    % \item
    $\bullet$
    \textbf{Bernoulli distributions} can be seen as an exponential family with $h(x) = 1$,
    and for a Bernoulli distribution of mean $\mu\in[0,1]$, denoted $B(\mu)$,
    the parameter is $\eta = \mu / (1 - \mu)$, giving $A(\eta) = \log(1 + \mathrm{e}^{\eta})$
    (with the limit behavior $\eta=+\infty$ if $\mu=0$).
    %
    Moreover, the KL divergence between $B(x)$ and $B(y)$, of parameters $\eta$, $\eta'$ is given by
    $\kl(x,y) = \KL_{B}(\eta,\eta') = x \log(x/y) + (1-x) \log((1-x)/(1-y))$,
    and it is also called the \emph{relative binary entropy}.
    It satisfies $\kl(x,y) \geq d_{1/4}(x,y)$, see Lemma~\ref{lem:2:Pinsker} in Appendix~\ref{proof:2:Pinsker}.

    $\bullet$
    \textbf{Gaussian distributions} of a fixed variance are a one-dimensional exponential family,
    and in this thesis we do not consider Gaussian with an unknown variance.
    For a variance of $\sigma^2$, the family uses
    % $T(x) = [x ; x^2]^T$ and
    $h(x) = 1/\sqrt{2\pi\sigma^2} \exp(-x^2/(2\sigma^2))$,
    and for a Gaussian distribution with mean $\mu\in\R$, denoted $\cN(\mu,\sigma^2)$,
    the parameter is $\eta = \mu/\sigma^2$, giving $A(\eta) = \mu^2/(2\sigma^2)$
    (see Chapter~8 of \cite{JordanCourseStatBerkeley}).
    % % For a generic variance,
    % This shows that the Gaussian distributions
    % % are a two-dimensional exponential families,
    % with a fixed variance $\sigma^2$ form indeed a one-dimensional exponential family.
    %
    Moreover, the KL divergence between $\cN(x,\sigma^2)$ and $\cN(x,\sigma^2)$, of parameters $\eta$, $\eta'$ is given by
    $d_{\sigma^2}(x,y) = \KL_{\cN,\sigma^2}(\eta,\eta') = (x-y)^2 / (2\sigma^2)$.
% \end{itemize}

@@

\begin{definition}[Regret]\label{def:2:regret}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
    For an algorithm $\cA$, a bandit problem of $K$ arms characterized by their means $\bm{\mu} = \mu_1,\dots,\mu_K$, with $\mu^* = \max_k \mu_k$, then its \emph{(mean) regret} at horizon $T$ is $R_T^{\cA}$ defined as
    %
    \begin{align}\label{eq:2:regret}
        R_T^{\cA}
        = \E\left[ \sum_{t=1}^T (Y_{k^*,t} - r(t)) \right]
        % = T \mu^* - \sum_{t=1}^T \E\left[ Y(t) \right]
        = T \mu^* - \sum_{t=1}^T \E\left[ r(t) \right].
    \end{align}
    \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar defnbar, comment if needed
\end{definition}

@@

\begin{lemma}[Regret decomposition]\label{lem:2:RegretDecomposition}
\begin{leftbar}[lemmabar]  % XXX leftbar lemmabar, comment if needed
    The (mean) regret $R_T^{\cA}$ can be decomposed as a sum of the number of selections of sub-optimal arms $k$, weighted by their gaps:
    \begin{align}\label{eq:2:RegretDecomposition}
        R_T^{\cA} = \sum_{k=1}^K \Delta_k \; \E[ N_k(T) ]
        = \sum_{\substack{k=1,\dots,K\\\Delta_k > 0}} \Delta_k \; \E[ N_k(T) ].
    \end{align}
    \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar lemmabar, comment if needed
\end{lemma}
%
\begin{proof}\label{proof:2:RegretDecomposition}
    We use the chain rule of expectation and a conditioning on $O_t$,
    because the expectation is taken on the randomness of the (\iid) samples $(Y_{k,t})_t$ and on the decisions of the player $(A(t))_t$, which are measurable wrt to the past observations $O_t = (U_0, Y_{A(1),1}, U_1, \dots, Y_{A(t-1),t-1}, U_{t-1})$.
    Thus we can rewrite the expected cumulated reward as follows:
    %
    % FIXME add a blank line or a negative space if it looks weird
    %
    % \vspace*{-20pt}
    \begin{align*}
        \E \left[ \sum_{t=1}^T r(t) \right]
        &= \E \left[ \sum_{t=1}^T \sum_{k=1}^K Y_{k,t} \mathbbm{1}(A(t) = k) \right]
        = \sum_{k=1}^K \sum_{t=1}^T \E \left[ Y_{k,t} \mathbbm{1}(A(t) = k) \right] \\
        &= \sum_{k=1}^K \sum_{t=1}^T \E \Bigl[ \E \left[ Y_{k,t} | O_t \right] \mathbbm{1}(A(t) = k) \Bigr]
        = \sum_{k=1}^K \sum_{t=1}^T \E \Bigl[ \E \left[ Y_{k,1} \right] \mathbbm{1}(A(t) = k) \Bigr] \\
        &= \sum_{k=1}^K \underbrace{\E \left[ Y_{k,1} \right]}_{\mu_k} \underbrace{\sum_{t=1}^T \E \left[ \mathbbm{1}(A(t) = k) \right]}_{= N_k(T)}
        % = \sum_{k=1}^K \mu_k \E \left[ \sum_{t=1}^T \mathbbm{1}(A(t) = k) \right]\\
        = \sum_{k=1}^K \mu_k \E \left[ N_k(T) \right].
    \end{align*}
    %
    Thus $R_T^{\cA} = T \mu^* - \sum_{t=1}^T \E\left[ r(t) \right] = T \mu^* - \sum_{k=1}^K \mu_k \E[ N_k(T) ] = \sum_{k=1}^K (\mu^* - \mu_k) \E[ N_k(T) ]$,
    as $\sum_{k=1}^K  \E[ N_k(T) ] = T$.
    The sum can then be simplified to only account for sub-optimal arms, \ie, arms $k$ such that $\Delta_k > 0$.
\end{proof}

\textbf{Consequences.}
%
Such decomposition of the regret can be useful for at least two reasons:

@@

\begin{theorem}\label{thm:2:secondLogTLowerBound}
\begin{leftbar}[theorembar]  % XXX leftbar theorembar, comment if needed
    If $\cA$ is uniformly efficient,
    then for any arbitrary problem instance $I$,
    there exists a constant $C_I$ depending only on $I$,
    and a time $T_0$ such that
    \begin{equation}
        \forall T \geq T_0, \;\;\; R_T^{\cA}(I) \geq C_I \log(T).
    \end{equation}
    \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar theorembar, comment if needed
\end{theorem}

@@

\paragraph{Worst-case lower-bound in $\Omega(\sqrt{T})$.}
%
For a fixed horizon $T$, it is interesting to note that one can find instances $I$ that are so ``hard'' that a logarithmic regret (lower or upper) bound that uses a constant $C_I$ no longer brings any information.
Indeed, we can naively bound the regret by $R_T^{\cA} \leq (\max_k \Delta_k) T$, and thus if $(\max_k \Delta_k)$ can be taken so small that $C_I \log(T) \gg (\max_k \Delta_k) T$, and thus a regret upper bound like $R_T^{\cA} \leq C_I \log(T)$ is useless.
%
\begin{proof}\label{proof:2:worstCaseLowerBound}
    For an example, consider any large horizon $T>10$, and $K=2$ Bernoulli arms.
    Chose any small $\varepsilon$ such that $\varepsilon \ll \sqrt{4 \log(T) / T}$, then chose let $\mu_1 = 1/2$ and $\mu_2 = 1/2 - \varepsilon$.
    The constant $C_I$ from the first case of Theorem~\ref{thm:2:forSecondLogTLowerBound} is $C_I = 1 / (4 \varepsilon)$ and $\Delta=\varepsilon$ and it becomes so large that
    the logarithmic upper-bound is useless in such case, as
    $\varepsilon \ll \sqrt{\frac{4 \log(T)}{T}} \Longleftrightarrow \frac{1}{4\varepsilon} \log(T) \gg \Delta T$.
\end{proof}

@@

On the other hand, if a lower-bound $d>0$ on the positive gaps is known beforehand, one can find a tuning of $T_0$ that gives a logarithmic regret.
%
To prove a bound on $p$, we use Hoeffding's inequality from \cite{hoeffding1963probability}, reminded below in Lemma~\ref{lem:2:HoeffdingInequality}.
%
We have $R_T \leq (\max_{k: \Delta_k > 0} \Delta_k) T_0 + p (T - T_0)$, thus Hoeffding's inequality gives $R_T \leq K T_0 + K \exp(-T_0 d^2/4) T$. Optimizing on $T_0$ gives $T_0 = K (4/d^2) \log(d^2 T / 4)$
and proves that the ``explore-then-exploit'' strategy can also obtain an order-optimal regret, $R_T \leq K (4/d^2) (1 + \log(d^2 T / 4)) = \bigO{K \log(T)}$, if it is tuned with a fixed time $T_0$ using prior knowledge on the problem (\ie, $d$) and the horizon $T$.
%
We note that this strategy can also achieve $R_T = \bigO{T^{2/3} (K \log(T))^{1/3}}$ without prior knowledge on the problem, as shown in Section~1 of \cite{Slivkins2019}.
%
\begin{proof}\label{proof:2:tuningExploreThenCommit}
    Let us prove the inequality $p \leq 2 K \exp(-T_0 d^2 / 4)$.\\
    \indent
    %
    First for the case of two arms:
    if the distributions are Bernoulli (or sub-Gaussian), we can use Hoeffding's inequality to control this probability $p$.
    %
    If $\mu_1 = \mu_2 + \Delta$, then
    $p = \bP(\widehat{\mu_1} < \widehat{\mu_2}) \leq \bP(\widehat{\mu_1} < \mu_1 - \Delta/2) + \bP(\widehat{\mu_2} > \mu_2 + \Delta/2)$, and both terms can be bounded by using Hoeffding's inequality with a (non-random) number of samples $n=T_0/2$,
    to obtain $p \leq 2 \exp(-T_0 \Delta^2 / 4)$.\\
    \indent
    %
    Then for $K\geq2$ arms, a simple union bound gives
    $p \leq 2 K \exp(-T_0 d^2 / 4)$, if $d$ is a lower-bound on the positive gaps $\Delta_k$.
    %
    In all cases, we showed that $p$ is controlled by $T_0$.
\end{proof}

\begin{lemma}[Hoeffding's inequality]\label{lem:2:HoeffdingInequality}
\begin{leftbar}[lemmabar]  % XXX leftbar lemmabar, comment if needed
    Let $Z_1,\dots,Z_n$ be $n$ \iid{} samples from a Bernoulli distribution of parameter $\mu$, of empirical mean $\widehat{Z_n} = \frac{1}{n} \sum\limits_{i=1}^n Z_i$, Hoeffding's inequality gives both
    \begin{align}\label{eq:2:HoeffdingInequality}
        \begin{cases}
            \forall x < \mu, \;\;\; & \bP(\widehat{Z_n} < x) \leq \exp(-2 n (x-\mu)^2),
            \\
            \forall x > \mu, \;\;\; & \bP(\widehat{Z_n} > x) \leq \exp(-2 n (x-\mu)^2).
        \end{cases}
    \end{align}
    \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar lemmabar, comment if needed
\end{lemma}

@@

\begin{theorem}[Regret bound for $\UCB_1$]\label{thm:2:UCBregretBound}
\begin{leftbar}[theorembar]  % XXX leftbar theorembar, comment if needed
    For simplicity, we restrict to $\alpha=2$.
    For any instance $I$ with $K$ arms with bounded distributions in $[0,1]$,
    of means $\mu_1,\dots,\mu_k$, and any horizon $T>K$,
    the $\UCB_1$ algorithm verifies the following finite-time regret bound
    \begin{equation}\label{eq:2:UCBregretBound}
        R_T^{\cA}(I) \leq \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T) + \sum_{k: \Delta_k>0} \Delta_k \left(1 + \frac{\pi^2}{3}\right)
        = \bigO{ \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T)}.
    \end{equation}
    \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar theorembar, comment if needed
\end{theorem}
%
\begin{proof}\label{proof:2:UCBregretBound}
    As we focus on bounded distributions in $[0,1]$, for which Hoeffding's inequality gives \eqref{eq:2:HoeffdingInequality}, as stated above in Lemma~\ref{lem:2:HoeffdingInequality}.
    Thanks to the regret decomposition from Lemma~\ref{lem:2:RegretDecomposition}, we can focus on bounding $\E[N_k(T)]$ for any sub-optimal arm $k$.\\
    %
    \indent
    The first trick consists in writing the following, which is true for any $n \geq 1$:
    \begin{equation}\label{eq:2:trickInUCBProof}
        N_k(T) = 1 + \sum_{t=K+1}^T \indic(A(t) = k)
        \leq n + \sum_{t=n}^T \indic(A(t) = k, N_k(t) \geq n).
    \end{equation}

@@

    More details on this proof can be found for instance in Section~1.3 of \cite{Slivkins2019}, while for instance Chapter~2 of \cite{Bubeck12} Theorem~2.1) and Chapter~7 of \cite{LattimoreBanditAlgorithmsBook} (Theorem~7.2) both give more general proofs, in particular for any value of $\alpha\geq\frac{1}{2}$.
\end{proof}

@@

\begin{proof}
    Fix $p\in[0,1]$, and consider the function $g(x) = \kl(p, p+x) - 2 x^2$ ($x$ will be $q-p$), for $x\in(-p, 1-p)$.
    We first observe that $g(0) = 0$, and we easily prove that $x=0$ is the unique minimizer of $g$ over its domain.
    By definition, $\kl$ is of class $\cC^1$ on its two variables, on $(0,1)$, so $g$ is of class $\cC^1$ on the interior of its domain, $(-p, 1-p)$.
    Let us compute its derivative, $g'(x) = (\frac{\partial \; \kl}{\partial 2})(p, p + x) - 4x$.
    Simple calculus yield
    \[ \left(\frac{\partial \; \kl}{\partial 2}\right) (p, p + x) = - \frac{p}{p+x} + \frac{1-p}{1-(p+x)} = -x \frac{(2(p + x) - 1)^2}{(p + x)(p + x - 1)}. \]
    % https://www.wolframalpha.com/input/?i=diff(p+*+log(p%2F(p%2Bx))+%2B+(1-p)+*+log((1-p)%2F(1-(p%2Bx)))+-+2*x**2,+x)
    We can verify that $g'(0) = -1 + 1 - 4 \times 0 = 0$, and it has the sign of $-x$ on its domain, as $p + x - 1 < 0$ because $x < 1 - p$, and $p + x > 0$.
    So $g'$ is first positive then negative, so $g$ is increasing then decreasing on its domain, with a change in its monotony at $x=0$, and thus it indeed has a unique minimizer.
    We conclude that $g(x) \geq g(0) = 0$ on $(-p, 1-p)$, and so $\kl(p,q) \geq 2 (p-q)^2$ if $x = q-p$.

    The edge case is the ends of the interval, that is if $x=-p$ or $x=1-p$, that is if $q=0$ or $q=1$.
    If $p\in(0,1)$, then computing the limit of $\kl(p,q)$ for $q\to0^+$ and $q\to1^-$ is obvious and gives $\kl(p,q) \to +\infty$, while the right-hand side of Pinsker's inequality stays finite (so the inequality is still true at the limit).
    If $p=0$ or $p=1$, the same computation works, by the convention that $p \log(p) = 0$ if $p=0$.
    %
    % Bonus: one can verify our computation in Python with the SymPy module:
    % \begin{verbatim}
    %     $ (bash) python
    %     >>> from sympy import *
    %     >>> x, p = symbols('x p')
    %     >>> g = p * log(p/(p+x)) + (1-p) * log((1-p)/(1-(p+x))) - 2*x**2
    %     >>> print(diff(g, x))
    %     -p/(p + x) - 4*x + (-p + 1)/(-p - x + 1)
    % \end{verbatim}
\end{proof}

% \TODOL{Write this ?}

>>>>

% DONE Je dois unifier les notations, et utiliser $Y_{k,t}$ pour les flux qui sont iid et $r(t)$ pour les récompenses effectivement observées (comme dans les chapitres 5 et 6)

@@

\begin{definition}[Kullback-Leibler divergence]\label{def:2:KLDivergence}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
    The KL divergence between two distributions $d_1$ and $d_2$ is defined as
    \begin{align}\label{eq:2:Kullback-LeiblerDivergenceExpFamily1}
        \KL\left(d_1, d_2\right) = \int d_1 \log\left(\frac{d_2}{d_1}\right) \lambda(\mathrm{d}x) = \E_{d_1} \left[ \log\left(\frac{d_2}{d_1}\right) \right] \in \mathbb{R}\cup\{\pm\infty\}.
    \end{align}
% \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar defnbar, comment if needed
\end{definition}

@@

\label{par:2:notationsExponentialFamiliesBernoulliGaussian}
\begin{itemize}
    \item
    % $\bullet$
    \textbf{Bernoulli distributions} can be seen as an exponential family with $h(x) = 1$,
    and for a Bernoulli distribution of mean $\mu\in[0,1]$, denoted $B(\mu)$,
    the parameter is $\eta = \mu / (1 - \mu)$, giving $A(\eta) = \log(1 + \mathrm{e}^{\eta})$
    (with the limit behavior $\eta=+\infty$ if $\mu=0$).

    The KL divergence between $B(x)$ and $B(y)$, of parameters $\eta$, $\eta'$ is given by
    $\kl(x,y) = \KL_{B}(\eta,\eta') = x \log(x/y) + (1-x) \log((1-x)/(1-y))$,
    and it is also called the \emph{relative binary entropy}.
    It satisfies $\kl(x,y) \geq d_{1/4}(x,y)$, see Lemma~\ref{lem:2:Pinsker} in Appendix~\ref{proof:2:Pinsker}.

    \item
    % $\bullet$
    \textbf{Gaussian distributions} of a fixed variance are a one-dimensional exponential family,
    and in this thesis we do not consider Gaussian with an unknown variance.
    For a variance of $\sigma^2$, the family uses
    % $T(x) = [x ; x^2]^T$ and
    $h(x) = 1/\sqrt{2\pi\sigma^2} \exp(-x^2/(2\sigma^2))$,
    and for a Gaussian distribution with mean $\mu\in\R$, denoted $\cN(\mu,\sigma^2)$,
    the parameter is $\eta = \mu/\sigma^2$, giving $A(\eta) = \mu^2/(2\sigma^2)$.
    % % For a generic variance,
    % This shows that the Gaussian distributions
    % % are a two-dimensional exponential families,
    % with a fixed variance $\sigma^2$ form indeed a one-dimensional exponential family.

    The KL divergence between $\cN(x,\sigma^2)$ and $\cN(x,\sigma^2)$, of parameters $\eta$, $\eta'$ is given by
    $d_{\sigma^2}(x,y) = \KL_{\cN,\sigma^2}(\eta,\eta') = (x-y)^2 / (2\sigma^2)$
    (see Chapter~8 of \cite{JordanCourseStatBerkeley}).
\end{itemize}

@@

\begin{definition}[Regret]\label{def:2:regret}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
    For an algorithm $\cA$, a bandit problem of $K$ arms characterized by their means $\bm{\mu} = \mu_1,\dots,\mu_K$, with $\mu^* = \max_k \mu_k$, then its \emph{(mean) regret} at horizon $T$ is $R_T^{\cA}$ defined as
    %
    \begin{align}\label{eq:2:regret}
        R_T^{\cA}
        = \E\left[ \sum_{t=1}^T (Y_{k^*,t} - r(t)) \right]
        % = T \mu^* - \sum_{t=1}^T \E\left[ Y(t) \right]
        = T \mu^* - \sum_{t=1}^T \E\left[ r(t) \right].
    \end{align}
% \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar defnbar, comment if needed
\end{definition}

@@

\begin{lemma}[Regret decomposition]\label{lem:2:RegretDecomposition}
\begin{leftbar}[lemmabar]  % XXX leftbar lemmabar, comment if needed
    The (mean) regret $R_T^{\cA}$ can be decomposed as a sum of the number of selections of sub-optimal arms $k$, weighted by their gaps:
    \begin{align}\label{eq:2:RegretDecomposition}
        R_T^{\cA} = \sum_{k=1}^K \Delta_k \; \E[ N_k(T) ]
        = \sum_{\substack{k=1,\dots,K\\\Delta_k > 0}} \Delta_k \; \E[ N_k(T) ].
    \end{align}
    % \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar lemmabar, comment if needed
\end{lemma}
%
\begin{smallproof}\label{proof:2:RegretDecomposition}
    We use the chain rule of expectation and a conditioning on $O_t$,
    because the expectation is taken on the randomness of the (\iid) samples $(Y_{k,t})_t$ and on the decisions of the player $(A(t))_t$, which are measurable wrt to the past observations $O_t = (U_0, Y_{A(1),1}, U_1, \dots, Y_{A(t-1),t-1}, U_{t-1})$.
    Thus we can rewrite the expected cumulated reward as follows:
    %
    % FIXME add a blank line or a negative space if it looks weird
    %
    % \vspace*{-20pt}
    \begin{align*}
        \E \left[ \sum_{t=1}^T r(t) \right]
        &= \E \left[ \sum_{t=1}^T \sum_{k=1}^K Y_{k,t} \mathbbm{1}(A(t) = k) \right]
        = \sum_{k=1}^K \sum_{t=1}^T \E \left[ Y_{k,t} \mathbbm{1}(A(t) = k) \right] \\
        &= \sum_{k=1}^K \sum_{t=1}^T \E \Bigl[ \E \left[ Y_{k,t} | O_t \right] \mathbbm{1}(A(t) = k) \Bigr]
        = \sum_{k=1}^K \sum_{t=1}^T \E \Bigl[ \E \left[ Y_{k,1} \right] \mathbbm{1}(A(t) = k) \Bigr] \\
        &= \sum_{k=1}^K \underbrace{\E \left[ Y_{k,1} \right]}_{\mu_k} \underbrace{\sum_{t=1}^T \E \left[ \mathbbm{1}(A(t) = k) \right]}_{= N_k(T)}
        % = \sum_{k=1}^K \mu_k \E \left[ \sum_{t=1}^T \mathbbm{1}(A(t) = k) \right]\\
        = \sum_{k=1}^K \mu_k \E \left[ N_k(T) \right].
    \end{align*}
    %
    Thus $R_T^{\cA} = T \mu^* - \sum_{t=1}^T \E\left[ r(t) \right] = T \mu^* - \sum_{k=1}^K \mu_k \E[ N_k(T) ] = \sum_{k=1}^K (\mu^* - \mu_k) \E[ N_k(T) ]$,
    as $\sum_{k=1}^K  \E[ N_k(T) ] = T$.
    The sum can then be simplified to only account for sub-optimal arms, \ie, arms $k$ such that $\Delta_k > 0$.
\end{smallproof}

\paragraph{Consequences.}
%
Such decomposition of the regret can be useful for at least two reasons:

@@

\begin{theorem}\label{thm:2:secondLogTLowerBound}
\begin{leftbar}[theorembar]  % XXX leftbar theorembar, comment if needed
    If $\cA$ is uniformly efficient,
    then for any arbitrary problem instance $I$,
    there exists a constant $C_I$ depending only on $I$,
    and a time $T_0$ such that
    \begin{equation}
        \forall T \geq T_0, \;\;\; R_T^{\cA}(I) \geq C_I \log(T).
    \end{equation}
    % \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar theorembar, comment if needed
\end{theorem}

@@

\paragraph{Worst-case lower-bound in $\Omega(\sqrt{T})$.}
%
For a fixed horizon $T$, it is interesting to note that one can find instances $I$ that are so ``hard'' that a logarithmic regret (lower or upper) bound that uses a constant $C_I$ no longer brings any information.
Indeed, we can naively bound the regret by $R_T^{\cA} \leq (\max_k \Delta_k) T$, and thus if $(\max_k \Delta_k)$ can be taken so small that $C_I \log(T) \gg (\max_k \Delta_k) T$, and thus a regret upper bound like $R_T^{\cA} \leq C_I \log(T)$ is useless.
%
\begin{smallproof}\label{proof:2:worstCaseLowerBound}
    For an example, consider any large horizon $T>10$, and $K=2$ Bernoulli arms.
    Chose any small $\varepsilon$ such that $\varepsilon \ll \sqrt{4 \log(T) / T}$, then chose let $\mu_1 = 1/2$ and $\mu_2 = 1/2 - \varepsilon$.
    The constant $C_I$ from the first case of Theorem~\ref{thm:2:forSecondLogTLowerBound} is $C_I = 1 / (4 \varepsilon)$ and $\Delta=\varepsilon$ and it becomes so large that
    the logarithmic upper-bound is useless in such case, as
    $\varepsilon \ll \sqrt{\frac{4 \log(T)}{T}} \Longleftrightarrow \frac{1}{4\varepsilon} \log(T) \gg \Delta T$.
\end{smallproof}

@@

On the other hand, if a lower-bound $d>0$ on the positive gaps is known beforehand, one can find a tuning of $T_0$ that gives a logarithmic regret.
%
To prove a bound on $p$, we use Hoeffding's inequality from \cite{hoeffding1963probability}, reminded below in Lemma~\ref{lem:2:HoeffdingInequality}.
%
We have $R_T \leq (\max_{k: \Delta_k > 0} \Delta_k) T_0 + p (T - T_0)$, thus Hoeffding's inequality gives $R_T \leq K T_0 + K \exp(-T_0 d^2/4) T$. Optimizing on $T_0$ gives $T_0 = K (4/d^2) \log(d^2 T / 4)$
and proves that the ``explore-then-exploit'' strategy can also obtain an order-optimal regret, $R_T \leq K (4/d^2) (1 + \log(d^2 T / 4)) = \bigO{K \log(T)}$, if it is tuned with a fixed time $T_0$ using prior knowledge on the problem (\ie, $d$) and the horizon $T$.
%
We note that this strategy can also achieve $R_T = \bigO{T^{2/3} (K \log(T))^{1/3}}$ without prior knowledge on the problem, as shown in Section~1 of \cite{Slivkins2019}.
%
\begin{smallproof}\label{proof:2:tuningExploreThenCommit}
    Let us prove the inequality $p \leq 2 K \exp(-T_0 d^2 / 4)$.\\
    \indent
    %
    First for the case of two arms:
    if the distributions are Bernoulli (or sub-Gaussian), we can use Hoeffding's inequality to control this probability $p$.
    %
    If $\mu_1 = \mu_2 + \Delta$, then
    $p = \bP(\widehat{\mu_1} < \widehat{\mu_2}) \leq \bP(\widehat{\mu_1} < \mu_1 - \Delta/2) + \bP(\widehat{\mu_2} > \mu_2 + \Delta/2)$, and both terms can be bounded by using Hoeffding's inequality with a (non-random) number of samples $n=T_0/2$,
    to obtain $p \leq 2 \exp(-T_0 \Delta^2 / 4)$.\\
    \indent
    %
    Then for $K\geq2$ arms, a simple union bound gives
    $p \leq 2 K \exp(-T_0 d^2 / 4)$, if $d$ is a lower-bound on the positive gaps $\Delta_k$.
    %
    In all cases, we showed that $p$ is controlled by $T_0$.
\end{smallproof}

\begin{lemma}[Hoeffding's inequality]\label{lem:2:HoeffdingInequality}
\begin{leftbar}[lemmabar]  % XXX leftbar lemmabar, comment if needed
    Let $Z_1,\dots,Z_n$ be $n$ \iid{} samples from a Bernoulli distribution of parameter $\mu$, of empirical mean $\widehat{Z_n} = \frac{1}{n} \sum\limits_{i=1}^n Z_i$, Hoeffding's inequality gives both
    \begin{align}\label{eq:2:HoeffdingInequality}
        \begin{cases}
            \forall x < \mu, \;\;\; & \bP(\widehat{Z_n} < x) \leq \exp(-2 n (x-\mu)^2),
            \\
            \forall x > \mu, \;\;\; & \bP(\widehat{Z_n} > x) \leq \exp(-2 n (x-\mu)^2).
        \end{cases}
    \end{align}
    % \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar lemmabar, comment if needed
\end{lemma}

@@

\begin{theorem}[Regret bound for $\UCB_1$]\label{thm:2:UCBregretBound}
\begin{leftbar}[theorembar]  % XXX leftbar theorembar, comment if needed
    For simplicity, we restrict to $\alpha=2$.
    For any instance $I$ with $K$ arms with bounded distributions in $[0,1]$,
    of means $\mu_1,\dots,\mu_k$, and any horizon $T>K$,
    the $\UCB_1$ algorithm verifies the following finite-time regret bound
    \begin{equation}\label{eq:2:UCBregretBound}
        R_T^{\cA}(I) \leq \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T) + \sum_{k: \Delta_k>0} \Delta_k \left(1 + \frac{\pi^2}{3}\right)
        = \bigO{ \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T)}.
    \end{equation}
    % \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar theorembar, comment if needed
\end{theorem}
%
\begin{smallproof}\label{proof:2:UCBregretBound}
    As we focus on bounded distributions in $[0,1]$, for which Hoeffding's inequality gives \eqref{eq:2:HoeffdingInequality}, as stated above in Lemma~\ref{lem:2:HoeffdingInequality}.
    Thanks to the regret decomposition from Lemma~\ref{lem:2:RegretDecomposition}, we can focus on bounding $\E[N_k(T)]$ for any sub-optimal arm $k$.\\
    %
    \indent
    The first trick consists in writing the following, which is true for any $n \geq 1$:
    \begin{equation}\label{eq:2:trickInUCBProof}
        N_k(T) = 1 + \sum_{t=K+1}^T \indic(A(t) = k)
        \leq n + \sum_{t=n}^T \indic(A(t) = k, N_k(t) \geq n).
    \end{equation}

@@

    More details on this proof can be found for instance in Section~1.3 of \cite{Slivkins2019}, while for instance Chapter~2 of \cite{Bubeck12} Theorem~2.1) and Chapter~7 of \cite{LattimoreBanditAlgorithmsBook} (Theorem~7.2) both give more general proofs, in particular for any value of $\alpha\geq\frac{1}{2}$.
\end{smallproof}

@@

\begin{smallproof}
    Fix $p\in[0,1]$, and consider the function $g(x) = \kl(p, p+x) - 2 x^2$ ($x$ will be $q-p$), for $x\in(-p, 1-p)$.
    We first observe that $g(0) = 0$, and we easily prove that $x=0$ is the unique minimizer of $g$ over its domain.
    By definition, $\kl$ is of class $\cC^1$ on its two variables, on $(0,1)$, so $g$ is of class $\cC^1$ on the interior of its domain, $(-p, 1-p)$.
    Let us compute its derivative, $g'(x) = (\frac{\partial \; \kl}{\partial 2})(p, p + x) - 4x$.
    Simple calculus yield
    \[ \left(\frac{\partial \; \kl}{\partial 2}\right) (p, p + x) = - \frac{p}{p+x} + \frac{1-p}{1-(p+x)} = -x \frac{(2(p + x) - 1)^2}{(p + x)(p + x - 1)}. \]
    % https://www.wolframalpha.com/input/?i=diff(p+*+log(p%2F(p%2Bx))+%2B+(1-p)+*+log((1-p)%2F(1-(p%2Bx)))+-+2*x**2,+x)
    We can verify that $g'(0) = -1 + 1 - 4 \times 0 = 0$, and it has the sign of $-x$ on its domain, as $p + x - 1 < 0$ because $x < 1 - p$, and $p + x > 0$.
    So $g'$ is first positive then negative, so $g$ is increasing then decreasing on its domain, with a change in its monotony at $x=0$, and thus it indeed has a unique minimizer.
    We conclude that $g(x) \geq g(0) = 0$ on $(-p, 1-p)$, and so $\kl(p,q) \geq 2 (p-q)^2$ if $x = q-p$.

    The edge case is the ends of the interval, that is if $x=-p$ or $x=1-p$, that is if $q=0$ or $q=1$.
    If $p\in(0,1)$, then computing the limit of $\kl(p,q)$ for $q\to0^+$ and $q\to1^-$ is obvious and gives $\kl(p,q) \to +\infty$, while the right-hand side of Pinsker's inequality stays finite (so the inequality is still true at the limit).
    If $p=0$ or $p=1$, the same computation works, by the convention that $p \log(p) = 0$ if $p=0$.
    %
    % Bonus: one can verify our computation in Python with the SymPy module:
    % \begin{verbatim}
    %     $ (bash) python
    %     >>> from sympy import *
    %     >>> x, p = symbols('x p')
    %     >>> g = p * log(p/(p+x)) + (1-p) * log((1-p)/(1-(p+x))) - 2*x**2
    %     >>> print(diff(g, x))
    %     -p/(p + x) - 4*x + (-p + 1)/(-p - x + 1)
    % \end{verbatim}
\end{smallproof}
